---
title: "Elevation and Species of U.S. National Parks"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    orientation: column
    vertical_layout: fill
    theme: sandstone
---

```{r setup, include=FALSE}
library(flexdashboard)
library(shiny)
library(leaflet)
library(plotly)
library(dplyr)
library(sf)
library(elevatr)
library(forcats)
library(tidyverse)
library(dplyr)
```

# Sidebar {.sidebar}

Use the **Explore metros** tab to explore neighborhood diversity for your chosen metropolitan area in 2020. The red line on the scatterplot represents a locally-weighted estimate of how diversity varies in the metropolitan area by distance from its urban core or cores. Click and drag on the scatterplot to highlight the corresponding Census tracts on the map below, and click on a Census tract on the map to generate a chart of race and ethnicity counts.

Click the **Compare over time** tab to examine how locally-weighted estimates of neighborhood diversity by distance from the urban core has varied between the 1990 and 2020 Censuses, and view maps of these shifts over time. To learn more about the project, click the **About** tab.

Application author: [Kyle Walker](https://walker-data.com)

Data sources: [NHGIS](https://nhgis.org), United States Census Bureau


# Explore National Parks

## Column {.column}

### Elevation of Top 15 Most Visited National Parks

```{r}

# Read in the National Park centroid data
np_centroid <- read_sf("../Raw_Data/NPS_Centroid/")

# Extract longitude and latitude from geometry
np_centroid <- np_centroid %>%
  mutate(
    lon = st_coordinates(geometry)[, 1],  # Longitude
    lat = st_coordinates(geometry)[, 2]  # Latitude
  )

# Read in most visited National Park data
most_visited_nps_species_data <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-10-08/most_visited_nps_species_data.csv') 

most_visited_nps_species_data <- most_visited_nps_species_data %>%
  rename(UNIT_NAME = ParkName) %>%
  distinct(UNIT_NAME, .keep_all = TRUE)

# Filter centroid data for top 15 National Parks

np_centroid <- np_centroid %>%
  filter(UNIT_NAME %in% most_visited_nps_species_data$UNIT_NAME)


# Get elevation data
elevation_data <- get_elev_point(np_centroid, prj = st_crs(np_centroid), src = "aws") %>%
  arrange(elevation)

# Reorder PARKNAME by elevation
elevation_data <- elevation_data %>%
  filter(UNIT_TYPE == "National Park") %>%
  mutate(PARKNAME = fct_reorder(PARKNAME, elevation))

# Selecting Great Smoky Mountains for highlighting
gsm <- elevation_data %>% 
  filter(PARKNAME == "Great Smoky Mountains")

# Interactive plot using Plotly
plot <- plot_ly(
  data = elevation_data,
  x = ~PARKNAME,
  y = ~elevation,
  type = 'scatter',
  mode = 'markers',
  text = ~paste0(
    "<b>Park Name:</b> ", PARKNAME, "<br>",
    "<b>Elevation:</b> ", elevation, " meters<br>",
    "<b>Latitude:</b> ", round(lat, 2), "<br>",
    "<b>Longitude:</b> ", round(lon, 2)
  ),
  hoverinfo = "text",
  marker = list(size = 10, color = "#92C54B")
) %>%
  layout(
    xaxis = list(title = "U.S. National Park Name", tickangle = 45),
    yaxis = list(title = "Elevation (meters)"),
    showlegend = FALSE
  )

# Add a special marker for the Great Smoky Mountains
plot <- plot %>%
  add_trace(
    data = gsm,
    x = ~PARKNAME,
    y = ~elevation,
    type = 'scatter',
    mode = 'markers+text',
    text = ~paste("Great Smoky Mountains<br> ", "Elevation: ", elevation, "m"),
    textposition = "left top",
    marker = list(size = 12, color = "black"),
    hoverinfo = "text"
  )

# Display the plot
plot

```



### Great Smoky Mountains National Park Map

```{r, message=FALSE, warning=FALSE}
# Load the necessary data
shapefile_path <- "../Raw_data/Great_Smoky_Mountains_National_Park_Boundary"
boundary <- st_read(shapefile_path, quiet = TRUE)

# transform the boundary into the correct file type for leaflet
boundary_wgs84 <- st_transform(boundary, crs = 4326) 


# Create a leaflet map with the Smokey Mountains boundary
leaflet(boundary_wgs84) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  addPolygons(
    fillColor = "lightblue",  # Fill color for the park boundary
    weight = 2,               # Line weight for the boundary
    color = "black",          # Border color of the boundary
    opacity = 1,              # Opacity of the boundary
    fillOpacity = 0.5         # Opacity of the fill
  )
# Render the map
map
```

## Column {.column}

### Species by Elevation in Great Smoky Mountains National Park {.content height="100%"}

```{r}
# Read in cleaned data
nps_species <- read_csv("../smoky_mountains.csv", show_col_types = FALSE)

prepare_data <- function(data) {
  # Exclude vascular plants
  data <- data %>%
    filter(TaxaGroup != "Vascular-plant") %>%
    filter(TaxaGroup != "Non-vascular plant") %>%
    filter(!grepl("Wide", ElevationRange))
  
  # Define the category levels
  category_levels <- c("Amphibian", "Bird", "Fungi", "Insect", "Mammal", "Slug/Snail", "Spider/Scorpion")
  
  # Convert CategoryName to factor with specified levels
  data <- data %>%
    mutate(CategoryName = factor(CategoryName, levels = category_levels))
  
  # Expand data for overlapping ranges
  expanded_data <- data %>%
    mutate(Elevation_Group = case_when(
      ElevationRange == "Low" ~ "Low",
      ElevationRange == "Mid" ~ "Mid",
      ElevationRange == "High" ~ "High",
      ElevationRange == "Low and High" ~ "Low",
      ElevationRange == "Low to High" ~ "Low",
      ElevationRange == "Low to Mid" ~ "Low",
      ElevationRange == "Mid to High" ~ "Mid"
    )) %>%
    bind_rows(
      data %>%
        filter(ElevationRange == "Low and High") %>%
        mutate(Elevation_Group = "High")
    ) %>%
    bind_rows(
      data %>%
        filter(ElevationRange == "Low to High") %>%
        mutate(Elevation_Group = "Mid")
    ) %>%
    bind_rows(
      data %>%
        filter(ElevationRange == "Low to High") %>%
        mutate(Elevation_Group = "High")
    ) %>%
    bind_rows(
      data %>%
        filter(ElevationRange == "Low to Mid") %>%
        mutate(Elevation_Group = "Mid")
    ) %>%
    bind_rows(
      data %>%
        filter(ElevationRange == "Mid to High") %>%
        mutate(Elevation_Group = "High")
    )
  
  return(expanded_data)
}


# Prepare the data
expanded_data <- prepare_data(nps_species)

# UI
ui <- fluidPage(
  # Remove the title panel, but keep the sidebar layout
  sidebarLayout(
    sidebarPanel(
      selectInput("elevation_group", "Select Elevation Group:", choices = c("Low", "Mid", "High"))
    ),
    mainPanel(
      plotOutput("species_plot")
    )
  )
)


# Server
server <- function(input, output) {
  # Reactive data filtering based on selected elevation group
  filtered_data <- reactive({
    expanded_data %>%
      filter(Elevation_Group == input$elevation_group) %>%
      group_by(CategoryName)
  })
  
  # New color palette
  new_colors <- c("#61D04F", "#2297E6", "#28E2E5", "#CD0BBC", "#CD0BBC", "#DF536B")
  
  # Create the plot
  output$species_plot <- renderPlot({
    ggplot(filtered_data(), aes(x = CategoryName, fill = Nativeness)) +
      geom_bar() +
      scale_y_continuous(expand = c(0,0),limits = c(0, 40))+
      scale_x_discrete(labels = c("Amphibian", "Bird", "Fungi", "Insect", 
                                  "Mammal", "Slug/Snail", "Spider/Scorpion"), 
                       drop = FALSE)+
      scale_fill_manual(values = new_colors)+
      geom_text(stat='count', aes(label=..count..),position = position_stack(vjust=0.5))+
      labs(
        title = paste(input$elevation_group, "Elevation Species"),
        y = "Count", 
        x = "Species Category"
      ) +
      theme_bw()+
      theme(
        axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 16),
        axis.title = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 20, face = "bold"),
        plot.subtitle = element_text(size = 15, face = "bold"),
        strip.text.x = element_text(size = 12, face = "bold"),
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 10)
      )
  })
}

# Run the application 
shinyApp(ui, server)
```


